Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-12-23 analysis/paper/London_HauloutBeahvior.Rmd
clear up text re: response variable
closes #19

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals. Our
goals were threefold. First, we wished to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

@@

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, the response variable was
the transmitted hourly percent-dry value.

>>>>

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

```{r binomialdata}

portion_10 <- function(d) {
  n <- nrow(d)
  f <- d %>% dplyr::filter(percent_dry <= 10) %>% 
    nrow()
  p10 <- f/n
  return(p10)
}

portion_90 <- function(d) {
  n <- nrow(d)
  f <- d %>% dplyr::filter(percent_dry >= 90) %>% 
    nrow()
  p90 <- f/n
  return(p90)
}

@@

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent dry.
